---
title: "Proportional Species Richness"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning = FALSE, message = FALSE)
```

```{r}

# Load all the required packages.

library(dplyr)
library(tidyr) # for spliting on the period see below
library(moments) # for calculating moments for skewness etc.
library(reshape2)
library(ggplot2)
library(corrplot)
library(knitr)

par(mfrow=c(1, 1)) 

# load the data on to the system.

setwd("/Users/kautilyareddy/Desktop/MA 334 Project")
Main_Proj_data <-  read.csv("proportional_species_richness_V3.csv") # you use V2 or V3

```

# Introduction

Bees, Macromoths, Ladybirds, Butterflies, Vascular_plants, Hoverflies, and Grasshoppers_._Crickets were the seven features we focused on during exploratory data analysis. Our goal was to find relations and trends that might be useful for further research by examining the possible connections between these characteristics. 
We were further provided with the data about Location, Eastings, Northings, DominantLandClass , ecological status, period (years between 1970 and 1990 named as period Y90 and span of 2000 to 2013 into period Y00).

## Exploratory Data Analysis (EDA)

The study found some intriguing results in our data analysis. The mean of the ecological status for all the taxonomy groups (BD11) to that of the mean of the seven features  (BD7) that were allocated were both almost similar.(See below plot) However, when we took the mean of each feature separately, we found a significant change with the Bees ( 0.58 )and Macromoths ( 0.86 ). Furthermore, the number of Ladybirds, Macromoths and butterflies was found to have a moderate positive correlation. This suggests that the when the Ladybirds population improves, butterflies and Macromoths population typically improves as well.  This holds true for Hoverflies, and Grasshoppers_._Crickets too. This link is significant because it implies that these species may need one another for survival. 
 	


```{r,fig.width=7,fig.height=4}

#DATA EXPLORATION

# selecting just the chosen 7 and the dom land class, eco measure and period

eco_selected <- c(2,5,7,9,10,11,12)   # a particular troublesome case
eco_selected_names <- names(Main_Proj_data)[eco_selected]

Var_demo <- Main_Proj_data%>%select(all_of(eco_selected),dominantLandClass,
                               ecologicalStatus,period)

mean_selected <- rowSums(Var_demo[1:7],na.rm=TRUE)/7 # mean the 7 columns, row wise

par(mfrow=c(1, 2))  # divide graph area in 2 columns
hist(mean_selected,col=heat.colors(n=10),main='Mean of BD11', xlab='Ecological Status');
hist(Var_demo$ecologicalStatus,col=heat.colors(n=10),main='Mean of BD7', xlab='Ecological Status')

# Correlation Analysis 
# Calculate the correlation matrix
corr <- cor(Main_Proj_data[c("Bees", "Macromoths", "Ladybirds", "Butterflies", "Vascular_plants", "Hoverflies", "Grasshoppers_._Crickets")])

Main_Proj_data$period <- as.factor(Main_Proj_data$period) # must set categorical vars
# Convert dominantLandClass to factor
Main_Proj_data$dominantLandClass <- as.factor(Main_Proj_data$dominantLandClass)

# We are going to use the wales data for further analyis.
Proj_data<- Main_Proj_data%>%filter(grepl("w",dominantLandClass))

# select  7 randomly chosen predictors to form the trial eco_stat
all <- c(2:11)
eco_selected <- sample(all,size=7, replace = FALSE)
eco_selected <- c(2,5,7,9,10,11,12)   # a particular troublesome case
eco_not_selected <- all[!(all%in%eco_selected)]
eco_names <- names(Proj_data[,2:12])
eco_selected_names <- names(Proj_data)[eco_selected]

# calculate the bio div measure over 7 taxinomic groups
mean_selected <- rowMeans(Proj_data[,eco_selected],na.rm=TRUE) # mean the 7 columns 
# add in the biodiversity measure which is the mean over 7 taxonomic groups
Proj_data_MA334 <- Proj_data%>%mutate(eco_status_7=mean_selected)

```


On the other hand, we took a sample of data for the wales region and there is a positive correlation (0.21) between the BD7 and the Easting, which meant that as we go towards the east the ecological status also tends to increase. However, towards the North we have a negative correlation (-0.11) which  suggest that as we go towards the North the ecology decreases.


```{r,fig.width=7,fig.height=4}

# extend data exploration; with correlations between continuous variables 
cont_vars <- Proj_data_MA334%>%select(c(eco_selected,13,14)) # includes easting and northing 
cormat <- round(x = cor(cont_vars,use="pairwise.complete.obs"), digits = 2)

# now use the eastings and northings (these may be better used as predictors )
par(mfrow=c(1,2))  # divide graph area in 2 columns
plot((Proj_data_MA334$eco_status_7~Proj_data_MA334$Easting),main = "Easting vs Ecological data_7",xlab = "Easting", ylab = "Ecolocical data")
plot((Proj_data_MA334$eco_status_7~Proj_data_MA334$Northing),main = "Northing vs Ecological data_7",xlab = "Northing", ylab = "Ecological data")  # for BD7


```

## Hypothesis Testing

Based on the results of the one-sample t-test, there is strong evidence to suggest that the true mean of the population (BD7) is not equal to 0. The calculated t-value was 5.5898 which is significant at the chosen significance level, and the 95 percent confidence interval does not include 0. This implies that there is a statistically significant difference between the sample mean and the hypothesized population mean of 0. 


```{r,fig.width=7,fig.height=4}

# HYPOTHESIS TESTING

# following code splits between the two periods to find the BD7 change
# however it may be better to use period as a predictor 

Proj_data_MA334_period <- Proj_data_MA334%>%select(Location,period,eco_status_7)
Proj_data_MA334_split <- Proj_data_MA334_period%>%pivot_wider(names_from =period,values_from=eco_status_7)
Proj_data_MA334_split <- Proj_data_MA334_split%>%mutate(BD7_change=Y00-Y70)

BD7_change <- Proj_data_MA334_split%>%pull(BD7_change)
t.test(BD7_change,mu=0)  # t test with H0: mu=0

```

We have also performed the Asymptotic two-sample Kolmogorov-Smirnov test, where we used the BD7 population and the BD11 population for their ecological status analysis. The text statistic (D) was 0.068441, and the p-value that was associated to it was 0.1701. The alternative hypothesis was two – sided and also the p-value being higher than the significance level led to no significant difference between both the distributions. 

```{r,fig.width=8,fig.height=4}

# comparing the two distributions of bio div based on 7 and 11 taxonomic groups 
ks.test(Proj_data_MA334$eco_status_7,Proj_data_MA334$ecologicalStatus)

par(mfrow=c(1, 2))  # divide graph area in 1 columns
qqplot(Proj_data_MA334$eco_status_7,Proj_data_MA334$ecologicalStatus,xlab = "Eco Status 7", ylab = "Ecological Status", main = "Eco distribution of BD11 and BD7")
abline(0,1,col="red")
# both cdfs together  and do a kolmogorov test H0: distributions are the same
BD7_cdf <- ecdf(Proj_data_MA334$eco_status_7)
BD11_cdf <- ecdf(Proj_data_MA334$ecologicalStatus)
plot(BD11_cdf,col="red", main = "Eco distribution for BD11 vs BD7")
lines(BD7_cdf,col="green")
        

```


## Simple linear regression 

A basic linear regression model with one predictor variable (Bd11) and one response variable (BD7) for the two different time periods in Wales and is displayed in summary. The explanatory variable's estimated Intercept and slope coefficient are displayed in the coefficient table. Both coefficients have very modest p-values (less than 0.001), suggesting they are likely to be non-zero. The R-squared score for Y70 was 0.9022 indicates that the predictor variable accounts for 90.22% of the variance in the responder variable. The residual standard error was 0.0307. On the other hand Y00 resulted in R-squared score of 0.9441 , indicating the predicator variable accounts for 94.41% of variance in the responder variable. The residual standard error was 0.0303 for Y00 which is quite similar tohat of the what it was for Y70. The coefficient was -0.17073 in the Y70 and  - 0.11068 in the Y00 which indicates a negative relation between the predictor variable and the response variable. There is substantial evidence that the predictor variable significantly contributes to explaining the variance in the response variable, as indicated by the F-huge statistic's value of 5.335e+04 and low p-value (0.001).

The above findings also suggests that the relationship was stable for a relatively good time and then gets negative over time. 


```{r,fig.width=5,fig.height=5}

# SIMPLE LINEAR REGGRESSION

Bd7 <- rowMeans(Proj_data[c("Bees", "Macromoths", "Ladybirds", "Butterflies", "Vascular_plants", "Hoverflies", "Grasshoppers_._Crickets")], na.rm = TRUE)
Bd11 <- rowMeans(Proj_data[c("Bees", "Bird", "Bryophytes", "Butterflies", "Carabids", "Hoverflies", "Isopods", "Ladybirds", "Macromoths", "Grasshoppers_._Crickets", "Vascular_plants")], na.rm = TRUE)
period = Proj_data$period

# create a dataframe with bd7 and bd11 columns
combined_df <- cbind(Bd7, Bd11 )

# Convert combined_df into a data frame
combined_df <- as.data.frame(combined_df)

#adding period from the main dataset
combined_df$period <- Proj_data$period

# Filter rows where period equals to Y70
df_filtered_Y70 <- subset(combined_df, period == "Y70")

# Perform simple linear regression between bd7 and bd11
model <- lm(Bd7 ~ Bd11, data = df_filtered_Y70)
summary(model)


# Filter rows where period equals to Y00
df_filtered_Y00 <- subset(combined_df, period == "Y00")

# Perform simple linear regression between bd7 and bd11
model_1 <- lm(Bd7 ~ Bd11, data = df_filtered_Y00)
summary(model_1)

```

The below plots gives us information 

```{r,fig.width=8,fig.height=7}

par(mfrow=c(2, 2))  # divide graph area in 1 columns

plot(jitter(fitted(model)),residuals(model),xlab="Fitted",ylab="Residuals",main = "Y70")
abline(h=0)
qqnorm(model$residuals, main = "Normal Q-Q Plot for Y70")
qqline(model$residuals,col="red")

plot(jitter(fitted(model_1)),residuals(model_1),xlab="Fitted",ylab="Residuals",main = "Y00")
abline(h=0)
qqnorm(model_1$residuals, main = "Normal Q-Q Plot for Y00")
qqline(model_1$residuals,col="red")


```

## Multiple Linear Regression 

Bees, Macromoths, Ladybirds, Butterflies, Vascular_plants, Hoverflies, and Grasshoppers_._Crickets are all included in the table of calculated coefficients. When all predictor variables are set to zero, the Intercept is 0.291024. 


```{r}

# MULTIPLE LINEAR REGGRESSION

# Multiple linear regression add graphs

BD4 <- rowMeans(Main_Proj_data[c('Carabids', 'Bryophytes', 'Isopods', "Bird")], na.rm =TRUE)
Main_Proj_data$BD4 <-BD4

#Creating a multi linear regression

```

When we have removed on significant variable, herein Butterflies, then the Ladybirds, Vascular_plants, Hoverflies, and Grasshoppers_._Crickets  all will have positive coefficients, indicating that rising populations of these organisms correlate with rising response variable levels. For instance, a one-unit increase in the Ladybirds population relates to a 0.089974 unit rise in the response variable. However, when the Macromoths and Bees population increase, the target variable declines, as seen by the negative coefficient.	

The multiple R-squared for the model is 0.5153, which indicates that the predictor variables account for 51.53 percent of the variance in the response variable. Taking into account the total number of predictors, the corrected R-squared value is 0.5148. With a p-value of 2.2e-16, the F-statistic is statistically significant, suggesting that one or more of the predictor variables significantly contributes to the model. Unaccounted-for volatility in the response variable is represented by the residual standard error (0.07173), which is the standard deviation of the residuals.

```{r}

#Creating a multi linear regression while dropping non-significant variables

model <- lm(BD4 ~ Bees + Macromoths + Ladybirds +  Vascular_plants + Hoverflies + Grasshoppers_._Crickets, data = Main_Proj_data)
# model summary
summary(model)

```

The report in ( figure 4 ) shows the outcomes of a stepwise selection procedure in linear regression using the Akaike Information Criterion (AIC) as the model selection criterion. The model first runs the entire model, which includes all predictor variables (Bees, Macromoths, Ladybirds, Butterflies, Vascular Plants, Hoverflies, Grasshoppers_.Crickets), and then goes through a stepwise selection procedure, deleting one variable at a time, until the AIC does not improve significantly. The output shows the changes in the AIC, sum of squares (sum of squares of residuals), and residual sum of squares (RSS) after removing each variable from the entire model. The AIC values show the quality of model fit, with lower values suggesting better fit. The findings show that all predictor variables are present in the optimal model for BD4 prediction (with the lowest AIC). 

```{r}

# Fit full model
full_model <- lm(BD4 ~ Bees + Macromoths + Ladybirds + Butterflies + Vascular_plants + Hoverflies + Grasshoppers_._Crickets, data = Main_Proj_data)

# Perform stepwise selection based on AIC
step_model <- step(full_model)


```

Macromoths, Bees, Butterflies, Grasshoppers_ Crickets, Ladybirds, Vascular Plants, and Hoverflies are arranged from least to most significant in terms of their contribution to model fit. 

## Open analysis

### Bees vs Period

After running a separate t-test to compare the mean bee populations of 'y00' and 'y70', we found evidence of a significant shift in the bee population. 

```{r,fig.width=8,fig.height=7}

## Open Analysis

# Perform independent t-test to compare the mean bee populations between two periods

t.test(Main_Proj_data$Bees ~ Main_Proj_data$period, var.equal = TRUE)

```

There was a statistically significant change in the bee population between the two time periods, as measured by a t-value of 25.465 and a p-value of less than 2.2e-16. The 95% confidence interval of (0.1896400, 0. 2212737) further supports the idea that the real mean difference is larger than zero. 
In the 'y00' period, the average number of bees was calculated to be 0.7077523, which is much greater than the 'y70' period average of 0.5022954. This points to a drastic decline in bee numbers between the two periods.


```{r,fig.width=2.3,fig.height=2.3}

#Analysis for occourance of Dominant land class
# Calculate the mean of 'Bees' based on categories present in 'Year'
mean_df <- aggregate(Bees~ period, data = Main_Proj_data, mean)

# Create a bar plot of the mean of 'Bees' based on categories present in 'Year'
ggplot(mean_df, aes(x = period, y = Bees)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal()+
  labs(title = "Mean of Bees by Year", x = "Year", y = "Mean of Bees")


```


To ensure the long-term viability of the bee population, this data is essential for environmental and conservationists to take the appropriate action.



```{r,fig.width=2.3,fig.height=2.5}

# Remove numbers from dominantLandClass
Main_Proj_data$dominantLandClass <- gsub("\\d+", "", Main_Proj_data$dominantLandClass)

plot_data <- Main_Proj_data%>%group_by(dominantLandClass)%>%count()%>%
  arrange(dominantLandClass)

ggplot(plot_data, aes(x = dominantLandClass, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal()+
  labs(title = "Land Occurrences", x = "Country", y = "Count") +
  scale_x_discrete(labels = c("e" = "England", "s" = "Scotland", "w" = "Wales"))


```

When we performed analysis furthermore on the Dominant land class, we grouped all the occurrences and found out that England had the most number of occurrences with about 3000, however Wales had the least with only around 500. This maybe because of the area of land available which we all know that the area of England was more compared to that of Scotland and Wales.  


```{r,fig.width=7,fig.height=4}

# Create a scatterplot of Bees vs. Butterflies with different colors for each variable
plot(cont_vars$Bees, cont_vars$Vascular_plants,
     xlab = "Bees", ylab = "Vascular Plants",
     main = "Scatterplot of Bees and Vascular Plants",
     col = ifelse(names(cont_vars) == "Bees", "green", "blue"))

# Add a legend to the plot
legend("topright", legend = c("Bees", "Vascular Plants"), pch = 1, col = c("green", "blue"))

```

We compared the Bees and Vascular Plants. The findings demonstrated that bees and Vascular Plants were related very well. The analysis showed strong evidence that wherever we found the existence of Vascular plants the existence of bees were precise too. This further tells us that we can find most of the bees population where we have vascular plants. (Figure 7)
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.




```{r}

table <- data.frame()
for(i in eco_selected){
  Mean<-round(mean(Proj_data_MA334[,i],na.rm = TRUE),digits = 2)
  Median<-round(median(Proj_data_MA334[,i],na.rm = TRUE),digits = 2)
  SD<-round(sd(Proj_data_MA334[,i],na.rm = TRUE),digits = 2)
  Skew<-round(skewness(Proj_data_MA334[,i],na.rm = TRUE),digits = 2)
  
  table <- rbind(table,
                 c(eco_names[i-1],Mean,Median,SD,Skew))
                 
                 }
colnames(table) <- c("taxi_group","Mean","Median","SD","Skewness")
tab<-table %>% select("taxi_group","Mean","Median","SD","Skewness")
kable(tab)


```

This table presents quantitative measures, including mean, median, and standard deviation, that have been calculated for the BD7 Taxonomic group. These measures provide insight into the ecological status of each species in the group. Additionally, ANOVA or regression analyses have been performed to examine any variances across the groups.

Below are the observation:
Macromoths exhibit a more symmetrical distribution with a skewness value close to zero compared to other groups.
Butterflies and Ladybirds have a negative mean distribution, suggesting that the distribution of species within these groups is more biased towards the negative end of the mean.


